---
title: "opentripplanner: getting started"
author: "Marcus Young, modified by Malcolm Morgan and Robin Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{opentripplanner-get-started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Introduction

```{r, eval=FALSE}
install.packages("devtools")
# remotes::install_github("ITSLeeds/opentripplanner")
devtools::install_local("opentripplanner-master.zip")

library(opentripplanner)
```

Before you can use the package, you should set-up the OpenTripPlanner Java program.
You will need to download the latest version of OTP from https://repo1.maven.org/maven2/org/opentripplanner/otp/ , which can be done as follows (this downloads the file into a folder called `programs` in your home directory on Linux/Mac):

```{r, eval=FALSE}
# download file (may want to do this manually)
path_otp <- "programs/otp.jar" # On Linux
```

## Building an OTP Graph

We will build an example graph for the Isle of Wight using some example data provided for the package. A graph is what OTP uses to find routes, and must be built out of the raw data provided. 


```{r eval=FALSE}
path_data <- getwd()

# na minha versao
log <- otp_build_graph(otp = path_otp, dir = path_data) 


```

This code will create a new file `Graph.obj` that will be saved in the location defined by `path_data`. 


## Launch OTP and load the graph

```{r}

system("java -Xmx2G -jar programs/otp.jar --router default --graphs graphs --server", wait = FALSE)


```

## Connecting to the OTP from R

Now you have the OTP running on your computer you can let R connect to the OTP.

```{r, eval = FALSE}
otpcon <- otp_connect()
```

## Getting a route from the OTP

Now we can use R to get a route from the OTP. OTP accepts latitude and longitude coordinates for a `fromPlace` (start of the journey) and `toPlace` (end of the journey).

```{r, eval = FALSE}
unifor <- c(-3.768882,-38.481988)
northshopping <- c(-3.734972,-38.56589)


route <- otp_plan(otpcon, fromPlace = unifor, toPlace = northshopping, mode = "BUS", date_time = as.POSIXct("2018-11-14 12:12:12"))
```

If you have the `tmap` package installed you can view the route using.

```{r, eval = FALSE}

mapview::mapview(route, zcol = "route_option")
```


## Stopping the OTP

As the OTP is running in JAVA it will conitune to run after you close R.

You can stop the OTP running using the command. **NOTE: This will stop all running JAVA applications!**

```{r, eval = FALSE}
otp_stop()
```

Congratulations, you now have your own multi-modal router planner!