---
title: "opentripplanner: getting started"
author: "Marcus Young, modified by Malcolm Morgan and Robin Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{opentripplanner-get-started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Introduction

This tutorial is an introduction to OpenTripPlanner (OTP) - an open-source and cross-platform multi-modal route planner written in JAVA. It uses imported Open Street Map (OSM) data for routing on the street and path network and supports
multi-agency public transport routing through imported GTFS feeds. It can also apply a digital elevation model to the OSM street network, allowing, for example, cycle-friendly routes to be requested. OTP has a web front-end that can be used by end-users and a sophisticated routing API.

A major advantage of running your own multi-modal route planner is the ability to carry out analysis using amended transport data. Services such as Google Maps or TransportAPI are based on current public transport schedules and the existing road network. OTP enables you to modify transit schedules and/or make changes to the underlying street network. By editing a local copy of OSM, you can model the effects of opening new roads, closing roads, or imposing other restrictions. You can also look back in time. For example, you might want to examine the effect of reductions in rural bus services on the accessibility of health facilities. To do this you would need a network with bus schedules as they were in previous years.

## Prerequisites

You will need to have installed R, RStudio, and Java 8.  If possible the 64 Bit version of Java is preferable, especially if you want to use OTP over large areas. If you are unfamiliar with R, Java, and Open Trip Planner see the Dummies Guide vignette included with this package.

## Installation

The following commands will install and attach the package:

```{r, eval=FALSE}
install.packages("devtools")
remotes::install_github("ITSLeeds/opentripplanner")
devtools::install_local("opentripplanner-master.zip")
library(opentripplanner)
```

Before you can use the package, you should set-up the OpenTripPlanner Java program.
You will need to download the latest version of OTP from https://repo1.maven.org/maven2/org/opentripplanner/otp/ , which can be done as follows (this downloads the file into a folder called `programs` in your home directory on Linux/Mac):

```{r, eval=FALSE}
# download file (may want to do this manually)
path_otp <- "otp/programs/otp.jar" # On Linux
# path_otp <- "C:/Users/Public/otp.jar" # On Windows

url_otp <- "https://repo1.maven.org/maven2/org/opentripplanner/otp/1.3.0/otp-1.3.0-shaded.jar"
download.file(url = url_otp, destfile = path_otp, mode="wb")
```

## Building an OTP Graph

We will build an example graph for the Isle of Wight using some example data provided for the package. A graph is what OTP uses to find routes, and must be built out of the raw data provided. 


```{r eval=FALSE}
# Path to the sample data
path_data <- "otp" # find the current working directory
# dir.create("graphs") # create a folder structure for the data
# dir.create("graphs/default")
# Download example data
# download.file("https://github.com/ITSLeeds/opentripplanner/releases/download/0.1/isle-of-wight-demo.zip", 
#               destfile = "isle-of-wight-demo.zip", mode="wb")
# unzip("isle-of-wight-demo.zip", exdir = "graphs/default")
# unlink("isle-of-wight-demo.zip")
# Run the OTP and Build a graph object
log <- otp_build_graph(otp = path_otp, dir = path_data) 

# na minha versao
log <- otp_build_graph(otp = path_otp, dir = path_data, router = "fortaleza") 


```

This code will create a new file `Graph.obj` that will be saved in the location defined by `path_data`. 

### Building your own OTP Graph

If you want to build your own graph for a different location follow these steps, and change your `path_data` variable to the folder with your data.

An OTP graph specifies every location in the region covered and how to travel between them, and is compiled by OTP using OSM data for the street and path network (used for walk, bicycle and drive modes) and GTFS data for transit scheduling.

Our first task is to create the folder and file structure expected by OTP. This is a base directory called `otp` which contains a sub-directory called `graphs`. Directories created under `graphs` are known as OTP routers and contain all the files required to build a graph. A single OTP instance can host several routers, for example covering different regions. 

Create a router called 'default', and include the GTFS and OSM files for the road network and default transit schedules for the railway on the Isle of Wight, along with `router-config.json`, as shown below:


```{r, engine='bash', eval=FALSE}
/ otp                         # Your top folder for storing all OTP data
  /graphs                     
     /default                 # Subfolder with the name or the router
         osm.pbf              # Required OSM road map
         router-config.json   # Required config file
         gtfs.zip             # Optional GTFS data
         dem.tif              # Optional Elevation data
         
```

`router-config.json` is actually read when the OTP server is started (in the next section), but we add it now for convenience.

## Launch OTP and load the graph

The next step is to start up your OTP server, running the router called 'default'. OTP will load the graph you created into memory and you will then be able to plan multi-modal routes using the web interface. From the `otp` directory, run the following command:

```{r, eval = FALSE}
otp_setup(otp = path_otp, dir = path_data)
```

```{r}

# system("java -Xmx2G -jar programs/otp.jar --router default --graphs graphs --server", wait = FALSE)


```

## Connecting to the OTP from R

Now you have the OTP running on your computer you can let R connect to the OTP.

```{r, eval = FALSE}
otpcon <- otp_connect()
```

## Getting a route from the OTP

Now we can use R to get a route from the OTP. OTP accepts latitude and longitude coordinates for a `fromPlace` (start of the journey) and `toPlace` (end of the journey).

```{r, eval = FALSE}
unifor <- c(-3.768882,-38.481988)
northshopping <- c(-3.734972,-38.56589)


route <- otp_plan(otpcon, fromPlace = unifor, toPlace = northshopping, date_time = as.POSIXct("2018-11-05"), 
                  mode = "TRANSIT")
```

## Stopping the OTP

As the OTP is running in JAVA it will conitune to run after you close R.

You can stop the OTP running using the command. **NOTE: This will stop all running JAVA applications!**

```{r, eval = FALSE}
otp_stop()
```

Congratulations, you now have your own multi-modal router planner!